### 2019-08-22 paper.tex
Initial Overleaf Import

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}

\begin{document}

\input{header}

\maketitle

\begin{abstract}

We present JuliaDB, a distributed database for analytical workloads
that is built solely in Julia, a high-level high-performance dynamic
language for numerical computing. JuliaDB is designed to allow seamless
use of user-defined Julia functions and data types. Julia's just-in-time compilation welds code from across Julia's package ecosystem as well as user code to generate efficient machine code allowing for high-performance data science in a coherent environment. We show example applications involving error analysis, mathematical optimization, online statistics, machine learning and linear algebra. 
\headingtable

\end{abstract}

\section{Introduction}

In~\cite{bezanson2017julia}, we describe the motivation behind Julia and how it was designed to address the two language problem: 
\begin{quote}
As long as the developers’ language is harder to
grasp than the users’ language, numerical computing will always be hindered. This is an essential part of the design philosophy of Julia: all basic functionality must be possible to implement in Julia—never force the programmer to resort to using C or Fortran.
\end{quote}

Modern data science workflows fall into two categories both of which show symptoms of the two language problem, the categories are:

\begin{enumerate}
\item Data is stored in traditional transaction-oriented databases (e.g. as Postgres and MySQL), queried with SQL, and analysed in high-level languages such as Python or R. This pairing suffers from two limitations: a) Most dialects of SQL, restrict custom data types and user defined functions, and b) The database server deals with on-disk data in a separate memory space from that of the analytics language making it an awkward combination for data science tasks.

\item Systems such as Pandas~\cite{mckinney-proc-scipy-2010} for Python and dplyr~\cite{dplyr} for R that provide a data storage and retrieval layer that is native to the host language. While they bring
the data into the memory space of the analytics language, they still
have some disadvantages: a) users must use vectorized database operations for performance since iterating over items in a loop can be expensive, and b) the need for vectorized operations to be 
implemented in C++, making it difficult to work with user defined data and functions. 
\end{enumerate}

Spark~\cite{spark} when used with Scala addresses many of the fundamental issues described above. However, Scala has been known to have performance concerns due to the design choices in the language~\cite{scalaslow}. In addition, users often use Spark through PySpark~\cite{pyspark} or SparkR~\cite{sparkr}, to gain the benefits of dynamically typed analytical languages. This brings us back to the two language problem.

We address the above mentioned two language problem in analytical databases with Julia~\cite{bezanson2017julia}. Its dynamic typing encourages exploratory and incremental programming. At the same time, users and database implementors have fine grained control over the memory layout of data. Struct types can be stored inline within arrays allowing efficient storage of user-defined
types. The just-in-time compiler can optimize data base operations and user defined functions as one system. Code from various packages written without the database in mind
can be used without performance penalties. In general, Julia finds the sweet-spot between dynamic
hassle-free languages such as Python and R helping the data scientist,
and efficient manually-memory-managed languages such as C or C++ helping
the database implementer greatly.

\section{Case study: manipulating measurements}
\label{sec:measurements}

Using the Measurements.jl ~\cite{giordano2016} package to represent physical measurements with some uncertainty. JuliaDB is unaware of the Measurement datatype, yet it can store measurements, read and write to disk and perform any database operation while using the math defined on the measurement type. In a two-language system, a performant implementation of the measurement datastructure must be done in a low-level language like C. For performant queries on databases containing measurements, one will have to implement an array of measurements datastructure in the low-level language as well. And tediously enumerate all possible array operations. But with JuliaDB and Julia, simply defining the measurement types and the primitive operations on Measurements is enough to achieve all the said features. Measurements can be stored efficiently in Julia arrays, operations such as `mean` and `sum` or broadcast on arrays of measurements fall out their existing generic implementations due to Julia's dynamic multiple-dispatch based JIT compiler.

Measurements.jl implements the ± operator
which represents an observation with some uncertainty associated
with it. Numeric operations on measuements propagate the
uncertainty as prescribed by linear error propagation theory:
For a function application $G(a,b,c...)$ where arguments are measurements $a=\overline{a} \pm \sigma_{a}, b=\overline{b} \pm \sigma_{b}, c=\overline{c} \pm \sigma_{c}$ the error of the output is found using the relationship:

$$
\begin{aligned}\sigma_{G}^{2}=&\left(\left.\frac{\partial G}{\partial a}\right|_{a=\overline{a}} \sigma_{a}\right)^{2}+\left(\left.\frac{\partial G}{\partial b}\right|_{b=\overline{b}} \sigma_{b}\right)^{2}+\left(\left.\frac{\partial G}{\partial c}\right|_{c=\overline{c}} \sigma_{c}\right)^{2}+\cdots \\
&+2\left(\frac{\partial G}{\partial a}\right)_{a=\overline{a}}\left(\frac{\partial G}{\partial b}\right)_{b=\overline{b}} \sigma_{a b}+2\left(\frac{\partial G}{\partial a}\right)_{a=\overline{a}}\left(\frac{\partial G}{\partial c}\right)_{c=\overline{c}} \sigma_{a c} \\ &+2\left(\frac{\partial G}{\partial b}\right)_{b=\overline{b}}\left(\frac{\partial G}{\partial c}\right)_{c=\overline{c}} \sigma_{b c}+\cdots \end{aligned}
$$\\

Where and $E[x]$ is the expected value of $x$.
Measurements.jl does not know about arrays, or JuliaDB, it just
defines how measurements get combined. Support for
measurements in the database in a non-julia analytical database
would involve having to write e.g. a mean function for an array
of measurements in C; in Julia, it emerges from dynamic
multiple dispatch, and compiles to C-like code.

Unlike a typical non-Julia analytical databases, without any
first class support for Measurements,
Pure-julia analytical database
We can store measurements directly in a JuliaDB table.

Perform groupby-mean without implementing the
mean function for an array of measurements in C.

Use OnlineStats for stats on measurements for lesser
allocation and parallelism.

\section{Designed of JuliaDB}

In the following subsection we describe the design of JuliaDB which empowers performance and interoperability.

\subsection{Memory layout}

The basic datastructure in JuliaDB is the table. A table can be constructed
from a Julia-native tuple or named tuple of column vectors. A table
acts as an iterator of rows. Rows are tuples or named tuples of items.

Although the interface to a table is as a row iterator, the data is
stored column-wise for fast uniform access. The columns can be accessed
using the \texttt{columns} function (returns a named tuple of column
vectors). The rows can be accessed as an array of structs using the
\texttt{rows} function rows is a struct of arrays view into the columnar
storage, unlike table, \texttt{rows} returns a legitimate Julia array.
Listing 1 shows the workings of rows and columns.
\begin{lstlisting}[language=Julia]
julia> t = table((a=[1,2,3], b=[2,3,4]))
Table with 3 rows, 2 columns:
a  b

1  2
2  3
3  4
julia> columns(t)
(a = [1, 2, 3], b = [2, 3, 4])
julia> rows(t)
3-element StructArray{NamedTuple{(:a, :b),Tuple{Int64,Int64}},1,...}:
(a = 1, b = 2)
(a = 2, b = 3)
(a = 3, b = 4) 
\end{lstlisting}

Tables have a standard set of operations, below we almost just mention
them, we defer more detailed specification to the API documentation
{[}cite{]}. Mainly, they support \texttt{map}, \texttt{reduce}, \texttt{groupreduce}, \texttt{groupby} and \texttt{join} operations.

\subsection{Sort-based indexing}

We implement groupby and join by traversing tables in the key-order. We use counting sort wherever possible (e.g. integer columns, PooledArrays) to do a small amount of initial work on the order of $O(k \log k)$ where $k$ is the number of unique values and then acheive a sorted permutation in $O n$ time where $n$ is the number of rows. In most systems where the implementation is done in C++, the ordering routines are restricted to a handful of primitive data types. In our system any user defined data type with an \code{isless} method to compare two items can take part in the sorting.

\textbf{Indexing: }For speeding up table operations, you can tell
JuliaDB to keep the rows of a table sorted by some of its fields.
These fields are called primary key fields. The primary key fields
can be of any data type which has an \code{isless} method associated
with it to compare two values. This includes but is not limited to
numbers, strings, symbols, dates, time periods, currencies, quantities
of comparable physical units (e.g. inch, mm).

\subsection{Distributed compute}

A distributed table is a distributed analogue of the one described
above. It is stored as many tables across multiple Julia processes,
\code{columns} returns a tuple of \emph{distributed arrays} and
\code{rows} returns a distributed array of tuples! The distributed
arrays implementation has the usual array operations (map, reduce,
hcat, vcat) defined on it in a distributed manner. Constructing a
table with one or more distributed columns using the table constructor
results in a distributed table. We use the word ``table'' from now on to refer interchangeably to both types of tables, unless
noted otherwise.

Here we show benchmarks of JuliaDB against other popular data science
databases. Where appropriate, we turned off JuliaDB's distributed
features for fair comparison.

\subsection{CSV loading}

\begin{figure}[h]
\centering \includegraphics[width=5in]{image8.png} \caption{Performance comparison for loading 815MB from the NYC taxi dataset}
\label{fig:nyctaxiload} 
\end{figure}
Our test for loading data is an 815 MB CSV file from the NYC Taxi
\& Limousine Commission’s records on every yellow cab trip in January
of 2017. Figure~\ref{fig:nyctaxiload} compares loading times of
this data where the date-time fields are parsed as time objects or
simply as strings.

\subsection{Querying Data}

Our tests for querying involve three different tasks: 
\begin{enumerate}
\item Get the average “fare amount” by type of vendor. This is a standard
“group by” operation, calculating a statistic (mean) grouped by the
unique values in another column (type of vendor). 
\item Get the distribution of “number of passengers” by “day of week”. Here
we wish to count the number of trips where the number of passengers
is one, two, three, etc. grouped by the day of the week. 
\item Get the distribution of “number of passengers” by whether the weekday
number is even (Monday, Wednesday, Friday) or odd (Sunday, Tuesday,
Thursday, Saturday). Here we perform a similar “group by” operation
to the previous task, but use an arbitrary “group by” variable that
needs to depend on a user-defined function. 
\end{enumerate}
JuliaDB is the only platform that does not have a slowdown associated
with a UDF, with pandas taking the biggest hit (Figure~\ref{fig:nyctaxiquery}).

\begin{figure}[h]
\centering \includegraphics[width=5in]{image4.png} \caption{Query times for 3 different query types across various systems}
\label{fig:nyctaxiquery} 
\end{figure}

\section{Future directions}

1. AxisArrays
2. Query planning
3. More support for unstructured data

\section{Acknowledgements}

Tanmay Mohapatra, David Anthoff, Jacob Quinn

\section{Additional Document Style Options}
\label{sec:additional_doc}
%
The following additional style option is available with the \verb juliacon  class file:
\vskip 6pt
Please place any additional command definitions at the very start of
the \LaTeX{} file, before the \verb \begin{document} . For example, user-defined
\verb \def  and \verb \newcommand   commands that define macros for
technical expressions should be placed here. Other author-defined
macros should be kept to a minimum.
\vskip 6pt
Commands that differ from the standard \LaTeX{} interface, or that
are provided in addition to the standard interface, are explained in
this guide. This guide is not a substitute for the \LaTeX{} manual itself.
Authors planning to submit their papers in \LaTeX{} are advised to use
\verb \juliacon.cls  as early as possible in the creation of their files.

%
%
%
%
\begin{table*}[t]
\tabcolsep22pt
\tbl{If necessary, the tables can be extended both columns.}{
\begin{tabular}{|l|l|c|c|}\hline
Label & \multicolumn{1}{c|}{Description}
& Number of Users &
Number of Queries\\\hline
Test 1 & Training Data &
\smash{\raise-7pt\hbox{70}} & 104\\
\cline{1-2}\cline{4-4}
Test 2 & Testing Data I & & 105\\\hline
Test 3 & Testing Data II & 30 & 119\\\hline
& Total & 100 & 328\\\hline
\end{tabular}}
\label{tab:symbols}
\begin{tabnote}
This is an example of table footnote.
\end{tabnote}
\end{table*}
% \begin{figure*}[t]
% \centerline{\includegraphics[width=11cm]{juliagraphs.png}}
% \caption{If necessary, the images can be extended both columns.}
%   \label{fig:sample_image}
% \end{figure*}

\section{Additional features}
\label{sec:additional_faci}
In addition to all the standard \LaTeX{} design elements, the \verb juliacon  class file includes the following features:
In general, once you have used the additional \verb juliacon.cls facilities
in your document, do not process it with a standard \LaTeX{} class
file.

\subsection{Titles, Author's Name, and Affiliation}
\label{subsub:title_auth}
The title of the article, author's name, and affiliation are used at the
beginning of the article (for the main title). These can be produced
using the following code:

\begin{verbatim}
\title{ This is an example of article title} }
\author{
   \large 1st Author \\[-3pt]
   \normalsize 1st author's affiliation  \\[-3pt]
    \normalsize 1st line of address \\[-3pt]
    \normalsize 2nd line of address \\[-3pt]
    \normalsize	1st author's email address \\[-3pt]
  \and
   \large 2nd Author \\[-3pt]
   \normalsize 2nd author's affiliation  \\[-3pt]
    \normalsize 1st line of address \\[-3pt]
    \normalsize 2nd line of address \\[-3pt]
    \normalsize	2nd author's email address \\[-3pt]
\and
   \large 3rd Author \\[-3pt]
   \normalsize 3rd author's affiliation  \\[-3pt]
    \normalsize 1st line of address \\[-3pt]
    \normalsize 2nd line of address \\[-3pt]
    \normalsize	3rd author's email address \\[-3pt]
}
\maketitle
\end{verbatim}

\subsection{Writing Julia code}

A special environment is already defined for Julia code,
built on top of \textit{listings} and \textit{jlcode}.

\begin{verbatim}
\begin{lstlisting}[language = Julia]
using Plots

x = -3.0:0.01:3.0
y = rand(length(x))
plot(x, y)
\end{lstlisting}
\end{verbatim}
\begin{lstlisting}[language = Julia]
using Plots

x = -3.0:0.01:3.0
y = rand(length(x))
plot(x, y)
\end{lstlisting}

\subsection{Abstracts, Key words, term etc...}
\label{subsub:abs_key_etc}

At the beginning of your article, the title should be generated
in the usual way using the \verb \maketitle  command. For genaral tem and keywords use
\verb \terms ,
\verb \keywords  commands respectively. The abstract should be enclosed
within an abstract environment, All these environment
can be produced using the following code:
\begin{verbatim}
\terms{Experimentation, Human Factors}

\keywords{Face animation, image-based modelling...}

\begin{abstract}
In this paper, we propose a new method for the
systematic determination of the model's base of
time varying delay system. This method based on
the construction of the classification data related
to the considered system. The number, the orders,
the time delay and the parameters of the local
models are generated automatically without any
knowledge about the full operating range of the
process. The parametric identification of the local
models is realized by a new recursive algorithm for
on line identification of systems with unknown time
delay. The proposed algorithm allows simultaneous
estimation of time delay and parameters of
discrete-time systems. The effectiveness of
the new method has been illustrated through
simulation.
\end{abstract}

\end{verbatim}

\section{Some guidelines}
\label{sec:some_guide}
The following notes may help you achieve the best effects with the
\verb juliacon  class file.

\subsection{Sections}
\label{subsub:sections}
\LaTeXe{}  provides four levels of section headings and they are all
defined in the \verb juliacon  class file:
\begin{itemize}
\item \verb \section   \item \verb \subsection  \item \verb \subsubsection  \item \verb \paragraph  \end{itemize}
Section headings are automatically converted to allcaps style.
\subsection{Lists}
\label{sec:lists}
%
The \verb juliacon   class file provides unnumbered lists using the
\verb unnumlist  environment for example,

\begin{unnumlist}
\item First unnumbered item which has no label and is indented from the
left margin.
\item Second unnumbered item.
\item Third unnumbered item.
\end{unnumlist}
The unnumbered list which has no label and is indented from the
left margin. was produced by:
\begin{verbatim}
\begin{unnumlist}
\item First unnumbered item...
\item Second unnumbered item...
\item Third unnumbered item...
\end{unnumlist}
\end{verbatim}

The \verb juliacon   class file also provides hyphen list using the
\verb itemize  environment for example,
\begin{itemize}
\item First unnumbered bulleted item which has no label and is indented
from the left margin.
\item Second unnumbered bulleted item.
\item Third unnumbered bulleted item which has no label and is indented
from the left margin.
\end{itemize}
was produced by:
\begin{verbatim}
\begin{itemize}
\item First item...
\item Second item...
\item Third item...
\end{itemize}
\end{verbatim}

Numbered list is also provided in acmtog class file using the
enumerate environment for example,
\begin{enumerate}
\item The attenuated and diluted stellar radiation.
\item Scattered radiation, and
\item Reradiation from other grains.
\end{enumerate}

was produced by:
\begin{verbatim}
\begin{enumerate}
\item The attenuated...
\item Scattered radiation, and...
\item Reradiation from other grains...
\end{enumerate}
\end{verbatim}
\subsection{Illustrations (or figures)}
\label{subsub:sec_Illus}
The \verb juliacon   class file will cope with most of the positioning of
your illustrations and you should not normally use the optional positional
qualifiers on the \verb figure   environment that would override
these decisions.
\vskip 6pt

%
\begin{figure}[t]
\centerline{\includegraphics[width=4cm]{juliagraphs.png}}
\caption{This is example of the image in a column.}
	\label{fig:sample_figure}
\end{figure}

The figure \ref{fig:sample_figure} is taken from the JuliaGraphs
organization \footnote{https://github.com/JuliaGraphs}.

Figure captions should be \emph{below} the figure itself, therefore the
\verb \caption  command should appear after the figure or space left for
an illustration. For example, Figure 1 is produced using the following
commands:

\begin{verbatim}
\begin{figure}
\centerline{\includegraphics[width=20pc]{Graphics.eps}}
\caption{An example of the testing process for a
binary tree. The globa null hypothesis is tested
first at level $\alpha$ (a), and the level of
individual variables is reached last (d). Note
that individual hypotheses can be tested at
level $\alpha/4$ and not $\alpha/8$ as one might
expect at first.}
\label{sample-figure_2}
\end{figure}
\end{verbatim}
Figures can be resized using first and second argument of
\verb \includegraphics   command. First argument is used for modifying
figure height and the second argument is used for modifying
figure width respectively.
\vskip 6pt
Cross-referencing of figures, tables, and numbered, displayed
equations using the \verb \label  and \verb \ref   commands is encouraged.
For example, in referencing Figure 1 above, we used
\verb Figure~\ref{sample-figure}   \subsection{Tables}
\label{subsub:sec_Tab}
The \verb juliacon   class file will cope with most of the positioning of
your tables and you should not normally use the optional positional qualifiers on the table environment which would override these
decisions. Table captions should be at the top.
\begin{verbatim}
\begin{table}
\tbl{Tuning Set and Testing Set}{
\begin{tabular}{|l|l|c|c|}\hline
Label & \multicolumn{1}{c|}{Description}
& Number of Users &
Number of Queries\\\hline
Train70 & Training Data &
\smash{\raise-7pt\hbox{70}} & 104\\
\cline{1-2}\cline{4-4}
Test70 & Testing Data I & & 105\\\hline
Test30 & Testing Data II & 30 & 119\\\hline
& Total & 100 & 328\\\hline
\end{tabular}}
\end{table}\end{verbatim}

\begin{table}
\tbl{Tuning Set and Testing Set}{
\begin{tabular}{|l|l|c|c|}\hline
Label & \multicolumn{1}{c|}{Description}
& Number of Users &
Number of Queries\\\hline
Test 1 & Training Data &
\smash{\raise-7pt\hbox{70}} & 104\\
\cline{1-2}\cline{4-4}
Test 2 & Testing Data I & & 105\\\hline
Test 3 & Testing Data II & 30 & 119\\\hline
& Total & 100 & 328\\\hline
\end{tabular}}
\end{table}
\subsection{Landscaping Pages}
\label{subsub:landscaping_pages}
If a table is too wide to fit the standard measure, it may be turned,
with its caption, to 90 degrees. Landscape tables cannot be produced
directly using the \verb juliacon   class file because \TeX{} itself cannot
turn the page, and not all device drivers provide such a facility.
The following procedure can be used to produce such pages.
\vskip 6pt
Use the package \verb rotating   in your document and change the coding
from
\begin{verbatim}
\begin{table}...\end{table}
to
\begin{sidewaystable}...\end{sidewaystable}
and for figures
\begin{figure}...\end{figure}
to
\begin{sidewaysfigure}...\end{sidewaysfigure}
\end{verbatim}

environments in your document to turn your table on the appropriate
page of your document. For instance, the following code prints
a page with the running head, a message half way down and the
table number towards the bottom.
\begin{verbatim}
\begin{sidewaystable}
\tbl{Landscape table caption to go here.}{...}
\label{landtab}
\end{sidewaystable}
\end{verbatim}

\subsection{Double Column Figure and Tables}
\label{subsub:double_fig_tab}
For generating the output of figures and tables in double column
we can use the following coding:

\begin{enumerate}
\item For Figures:
\begin{verbatim}
\begin{figure*}...\end{figure*}
\end{verbatim}
\item For landscape figures:
\begin{verbatim}
\begin{sidewaysfigure*}...\end{sidewaysfigure*}
\end{verbatim}
\item For Tables:
\begin{verbatim}
\begin{table*}...\end{table*}
\end{verbatim}
\item For landscape tables:
\begin{verbatim}
\begin{sidewaystable*}...\end{sidewaystable*}
\end{verbatim}
\end{enumerate}

\subsection{Typesetting Mathematics}
\label{subsub:type_math}
The \verb juliacon class file will set displayed mathematics with center to
the column width, provided that you use the \LaTeXe{} standard of
open and closed square brackets as delimiters.
The equation
\[
\sum_{i=1}^p \lambda_i = (S)
\]

was typeset using the acmtog class file with the commands

\begin{verbatim}
\[
\sum_{i=1}^p \lambda_i = (S)
\]
\end{verbatim}

For display equations, cross-referencing is encouraged. For example,
\begin{verbatim}
\begin{equation}
(n-1)^{-1} \sum^n_{i=1} (X_i - \overline{X})^2.
\label{eq:samplevar}
\end{equation}
Equation~(\ref{eq:samplevar}) gives the formula for
sample variance.
\end{verbatim}
The following output is generated with the above coding:
\begin{equation}
(n-1)^{-1} \sum^n_{i=1} (X_i - \overline{X})^2.
\label{eq:samplevar}
\end{equation}
Equation~(\ref{eq:samplevar}) gives the formula for
sample variance.

\subsection{Enunciations}
\label{subsub:enunciation}
The \verb juliacon   class file generates the enunciations with the help of
the following commands:
\begin{verbatim}
\begin{theorem}...\end{theorem}
\begin{strategy}...\end{strategy}
\begin{property}...\end{property}
\begin{proposition}...\end{proposition}
\begin{lemma}...\end{lemma}
\begin{example}...\end{example}
\begin{proof}...\end{proof}
\begin{definition}...\end{definition}
\begin{algorithm}...\end{algorithm}
\begin{remark}...\end{remark}
\end{verbatim}
The above-mentioned coding can also include optional arguments
such as
\begin{verbatim}
\begin{theorem}[...]. Example for theorem:
\begin{theorem}[Generalized Poincare Conjecture]
Four score and seven ... created equal.
\end{theorem}
\end{verbatim}

\begin{theorem}[Generalized Poincare Conjecture]
Four score and seven years ago our fathers brought forth,
upon this continent, a new nation, conceived in Liberty,
 and dedicated to the proposition that all men are
created equal.
\end{theorem}

\subsection{Extract}
\label{subsub:extract}
Extract environment should be coded within
\begin{verbatim}
\begin{extract}..\end{extract}
\end{verbatim}

\subsection{Balancing column at last page}
\label{subsub:Balance}
For balancing the both column length at last page use :
\begin{verbatim}
\vadjust{\vfill\pagebreak}
\end{verbatim}

%\vadjust{\vfill\pagebreak}

at appropriate place in your \TeX{} file or in bibliography file.

\section{Handling references}
\label{subsub:references}
References are most easily (and correctly) generated using the
BIBTEX, which is easily invoked via
\begin{verbatim}
\bibliographystyle{juliacon}
\bibliography{ref}
\end{verbatim}
When submitting the document source (.tex) file to external
parties, the ref.bib file should be sent with it.
\cite{bezanson2017julia}

\input{bib.tex}

\end{document}

% Inspired by the International Journal of Computer Applications template

### 2019-08-23 paper.tex
more design

@@
 
-\section{Designed of JuliaDB}
+\section{Design of JuliaDB}
 
@@
 
-\subsection{Memory layout}
+\subsection{Columnar memory layout}
+The basic datastructure in JuliaDB is the table.
 
-The basic datastructure in JuliaDB is the table. A table can be constructed
-from a Julia-native tuple or named tuple of column vectors. A table
-acts as an iterator of rows. Rows are tuples or named tuples of items.
-
-Although the interface to a table is as a row iterator, the data is
-stored column-wise for fast uniform access. The columns can be accessed
-using the \texttt{columns} function (returns a named tuple of column
-vectors). The rows can be accessed as an array of structs using the
-\texttt{rows} function rows is a struct of arrays view into the columnar
-storage, unlike table, \texttt{rows} returns a legitimate Julia array.
-Listing 1 shows the workings of rows and columns.
-\begin{lstlisting}[language=Julia]
-julia> t = table((a=[1,2,3], b=[2,3,4]))
-Table with 3 rows, 2 columns:
-a  b
-
-1  2
-2  3
-3  4
-julia> columns(t)
-(a = [1, 2, 3], b = [2, 3, 4])
-julia> rows(t)
-3-element StructArray{NamedTuple{(:a, :b),Tuple{Int64,Int64}},1,...}:
-(a = 1, b = 2)
-(a = 2, b = 3)
-(a = 3, b = 4) 
+\begin{lstlisting}[language=Julia,caption=A table]
+lotr = table((name=StringArray(["Frodo", "Samwise", "Bilbo", "Aragorn",
+                                "Boromir", "Gimli", "Elrond", "Legolas"]),
+             race=PooledArray(["Hobbit", "Hobbit", "Hobbit", "Human",
+                               "Human", "Dwarf", "Elf", "Elf"]),
+             age=[33, 61, 111, 87, 41, 139, 6000, 2931]))
 \end{lstlisting}
 
-Tables have a standard set of operations, below we almost just mention
-them, we defer more detailed specification to the API documentation
-{[}cite{]}. Mainly, they support \texttt{map}, \texttt{reduce}, \texttt{groupreduce}, \texttt{groupby} and \texttt{join} operations.
+Listing 1 shows a table constructed from a named tuple of column vectors.
+Although a table stores data as a named tuple of columns, it acts as an iterator of rows.
+Rows are tuples or named tuples of items.
+
+The columns can be accessed
+using the \texttt{columns} function (e.g. \texttt{rows(lotr)} returns a named tuple of column
+vectors provided to \texttt{table} during construction) while the rows can be accessed as an array of tuples using the
+\texttt{rows} function rows is a struct of arrays view into the columnar
+storage, unlike \texttt{table} which returns a JuliaDB table, \texttt{rows} returns a legitimate Julia array.
+
+Tables support \texttt{map}, \texttt{reduce}, \texttt{groupreduce}, \texttt{groupby} and \texttt{join} operations.
+
+Listing 1 also illustrates use of \texttt{StringArray} and \texttt{PooledArray} array types as columns. These special array types enable efficient packing of String and categorical data respectively.
+
+A \texttt{StringArray}'s memory layout is showin in Figure \ref{fig:stringarray}.
+The contents of strings (unicode character codes) are packed into a buffer of 8-bit unsigend integers.
+\texttt{StringArray} also stores the lengths of each string and the index (offset) at which each of them starts.
+Although it would suffice to store just the length of each string, storing offset also allows us to implement efficient mutation operations.
+Setting an element of the array appends the new string to the end of the buffer and replaces the length and offset with the corresponding new length and offsets for that item.
+Operations such as \texttt{sort!} and \texttt{permute!} (in place permutation of items) can simply modify the length and offset arrays and leave the buffer unmodified.
+Wherever suitable, in code internal to JuliaDB, we create \texttt{WeakRefString} instead of creating String objects when we index into a \texttt{StringArray}, this greatly reduces GC-load when dealing with string columns in operations such as groupby and join.
+
+A \texttt{PooledArray}'s memory layout is shown in Figure \ref{fig:pooledarray}.
+It is useful in compressing data with low number of unique items as compared to the total number of items.
+PooledArrays stores a dictionary mapping each unique item to a \texttt{ref} integer, and then an array of refs which tells which unique element appears in what position in the uncompressed array.
 

### 2019-08-23 paper.tex
Updates from Overleaf

--- 2019-08-22 paper.tex
@@
 Using the Measurements.jl ~\cite{giordano2016} package to represent physical measurements with some uncertainty.
-JuliaDB is unaware of the Measurement datatype, yet it can store measurements, read and write to disk and perform any database operation while using the math defined on the measurement type.
+JuliaDB is unaware of the Measurement datatype, yet it can store measurements, read and write to disk, and perform any database operation while using the math defined on the measurement type.
 In a two-language system, a performant implementation of the measurement datastructure must be done in a low-level language like C.
-For performant queries on databases containing measurements, one will have to implement an array of measurements datastructure in the low-level language as well.
-And tediously enumerate all possible array operations.
+For performant queries on columns containing measurements, one will have to implement an array of measurements datastructure in the low-level language as well that supports many array operations.
 But with JuliaDB and Julia, simply defining the measurement types and the primitive operations on Measurements is enough to achieve all the said features.

### 2019-08-28 paper.tex
fig

@@
 
-We present JuliaDB, a distributed database for analytical workloads
-that is built solely in Julia, a high-level high-performance dynamic
-language for numerical computing. JuliaDB is designed to allow seamless
-use of user-defined Julia functions and data types.
+We present JuliaDB, a distributed database for analytical workloads that is built solely in Julia, a high-level high-performance dynamic language for numerical computing.
+JuliaDB is designed to allow seamless use of user-defined Julia functions and data types.
 Julia's just-in-time compilation welds code from across Julia's package ecosystem as well as user code to generate efficient machine code allowing for high-performance data science in a coherent environment.
-We show example applications involving error analysis, mathematical optimization, online statistics, machine learning and linear algebra.
+We show example applications involving error analysis, and online statistics.
+Our benchmark shows that the system can perform on-par with or betters that of pandas and tidyverse in R.
 \headingtable
@@
 
-\begin{lstlisting}[language=Julia,caption=A table]
-lotr = table((name=StringArray(["Frodo", "Samwise", "Bilbo", "Aragorn",
-                                "Boromir", "Gimli", "Elrond", "Legolas"]),
-             race=PooledArray(["Hobbit", "Hobbit", "Hobbit", "Human",
-                               "Human", "Dwarf", "Elf", "Elf"]),
-             age=[33, 61, 111, 87, 41, 139, 6000, 2931]))
+\begin{lstlisting}[language=Julia]
+lotr = table((name=StringArray(["Frodo", "Samwise", "Bilbo",
+                                "Aragorn", "Boromir", "Gimli",
+                                "Elrond", "Legolas"]),
+             race=PooledArray(["Hobbit", "Hobbit", "Hobbit",
+                               "Human", "Human", "Dwarf",
+                               "Elf", "Elf"]),
+             age=[33, 61, 111, 87, 41, 139, 6000, 2931]));
+
+collect(lotr)
+
+#=
+8-element Array{NamedTuple{(:name, :race, :age),
+                Tuple{String,String,Int64}},1}:
+ (name = "Frodo", race = "Hobbit", age = 33)  
+ (name = "Samwise", race = "Hobbit", age = 61)
+ (name = "Bilbo", race = "Hobbit", age = 111) 
+ (name = "Aragorn", race = "Human", age = 87) 
+ (name = "Boromir", race = "Human", age = 41) 
+ (name = "Gimli", race = "Dwarf", age = 139)  
+ (name = "Elrond", race = "Elf", age = 6000)  
+ (name = "Legolas", race = "Elf", age = 2931) 
+=#
 \end{lstlisting}
+Listing 1. \emph{A table stores a named tuple of columns, but produces named tuple rows when iterated.}\\
 
@@
 
-The columns can be accessed
-using the \texttt{columns} function (e.g. \texttt{rows(lotr)} returns a named tuple of column
-vectors provided to \texttt{table} during construction) while the rows can be accessed as an array of tuples using the
-\texttt{rows} function rows is a struct of arrays view into the columnar
-storage, unlike \texttt{table} which returns a JuliaDB table, \texttt{rows} returns a legitimate Julia array.
+The columns can be accessed using the \texttt{columns} function (e.g.
+\texttt{rows(lotr)} returns a named tuple of column vectors provided to \texttt{table} during construction) while the rows can be accessed as an array of tuples using the \texttt{rows} function rows is a struct of arrays view into the columnar storage, unlike \texttt{table} which returns a JuliaDB table, \texttt{rows} returns a legitimate Julia array.
 
@@
 
-A \texttt{StringArray}'s memory layout is showin in Figure \ref{fig:stringarray}.
+\begin{figure*}[t]
+\centerline{\includegraphics[width=11cm]{stringarray.png}}
+\caption{StringArray layout.}
+  \label{fig:stringarray}
+\end{figure*}
+
+A \texttt{StringArray}'s memory layout is show in in Figure \ref{fig:stringarray}.
 The contents of strings (unicode character codes) are packed into a buffer of 8-bit unsigend integers.
@@
 
-\section{Additional Document Style Options}
-\label{sec:additional_doc}
-%
-The following additional style option is available with the \verb juliacon  class file:
-\vskip 6pt
-Please place any additional command definitions at the very start of
-the \LaTeX{} file, before the \verb \begin{document} . For example, user-defined
-\verb \def  and \verb \newcommand   commands that define macros for
-technical expressions should be placed here. Other author-defined
-macros should be kept to a minimum.
-\vskip 6pt
-Commands that differ from the standard \LaTeX{} interface, or that
-are provided in addition to the standard interface, are explained in
-this guide. This guide is not a substitute for the \LaTeX{} manual itself.
-Authors planning to submit their papers in \LaTeX{} are advised to use
-\verb \juliacon.cls  as early as possible in the creation of their files.
-
-%
-%
-%
-%
-\begin{table*}[t]
-\tabcolsep22pt
-\tbl{If necessary, the tables can be extended both columns.}{
-\begin{tabular}{|l|l|c|c|}\hline
-Label & \multicolumn{1}{c|}{Description}
-& Number of Users &
-Number of Queries\\\hline
-Test 1 & Training Data &
-\smash{\raise-7pt\hbox{70}} & 104\\
-\cline{1-2}\cline{4-4}
-Test 2 & Testing Data I & & 105\\\hline
-Test 3 & Testing Data II & 30 & 119\\\hline
-& Total & 100 & 328\\\hline
-\end{tabular}}
-\label{tab:symbols}
-\begin{tabnote}
-This is an example of table footnote.
-\end{tabnote}
-\end{table*}
-% \begin{figure*}[t]
-% \centerline{\includegraphics[width=11cm]{juliagraphs.png}}
-% \caption{If necessary, the images can be extended both columns.}
-%   \label{fig:sample_image}
-% \end{figure*}
-
-\section{Additional features}
-\label{sec:additional_faci}
-In addition to all the standard \LaTeX{} design elements, the \verb juliacon  class file includes the following features:
-In general, once you have used the additional \verb juliacon.cls facilities
-in your document, do not process it with a standard \LaTeX{} class
-file.
-
-\subsection{Titles, Author's Name, and Affiliation}
-\label{subsub:title_auth}
-The title of the article, author's name, and affiliation are used at the
-beginning of the article (for the main title). These can be produced
-using the following code:
-
-\begin{verbatim}
-\title{ This is an example of article title} }
-\author{
-   \large 1st Author \\[-3pt]
-   \normalsize 1st author's affiliation  \\[-3pt]
-    \normalsize 1st line of address \\[-3pt]
-    \normalsize 2nd line of address \\[-3pt]
-    \normalsize	1st author's email address \\[-3pt]
-  \and
-   \large 2nd Author \\[-3pt]
-   \normalsize 2nd author's affiliation  \\[-3pt]
-    \normalsize 1st line of address \\[-3pt]
-    \normalsize 2nd line of address \\[-3pt]
-    \normalsize	2nd author's email address \\[-3pt]
-\and
-   \large 3rd Author \\[-3pt]
-   \normalsize 3rd author's affiliation  \\[-3pt]
-    \normalsize 1st line of address \\[-3pt]
-    \normalsize 2nd line of address \\[-3pt]
-    \normalsize	3rd author's email address \\[-3pt]
-}
-\maketitle
-\end{verbatim}
-
-\subsection{Writing Julia code}
-
-A special environment is already defined for Julia code,
-built on top of \textit{listings} and \textit{jlcode}.
-
-\begin{verbatim}
-\begin{lstlisting}[language = Julia]
-using Plots
-
-x = -3.0:0.01:3.0
-y = rand(length(x))
-plot(x, y)
-\end{lstlisting}
-\end{verbatim}
-\begin{lstlisting}[language = Julia]
-using Plots
-
-x = -3.0:0.01:3.0
-y = rand(length(x))
-plot(x, y)
-\end{lstlisting}
-
-\subsection{Abstracts, Key words, term etc...}
-\label{subsub:abs_key_etc}
-
-At the beginning of your article, the title should be generated
-in the usual way using the \verb \maketitle  command. For genaral tem and keywords use
-\verb \terms ,
-\verb \keywords  commands respectively. The abstract should be enclosed
-within an abstract environment, All these environment
-can be produced using the following code:
-\begin{verbatim}
-\terms{Experimentation, Human Factors}
-
-\keywords{Face animation, image-based modelling...}
-
-\begin{abstract}
-In this paper, we propose a new method for the
-systematic determination of the model's base of
-time varying delay system. This method based on
-the construction of the classification data related
-to the considered system. The number, the orders,
-the time delay and the parameters of the local
-models are generated automatically without any
-knowledge about the full operating range of the
-process. The parametric identification of the local
-models is realized by a new recursive algorithm for
-on line identification of systems with unknown time
-delay. The proposed algorithm allows simultaneous
-estimation of time delay and parameters of
-discrete-time systems. The effectiveness of
-the new method has been illustrated through
-simulation.
-\end{abstract}
-
-\end{verbatim}
-
-\section{Some guidelines}
-\label{sec:some_guide}
-The following notes may help you achieve the best effects with the
-\verb juliacon  class file.
-
-\subsection{Sections}
-\label{subsub:sections}
-\LaTeXe{}  provides four levels of section headings and they are all
-defined in the \verb juliacon  class file:
-\begin{itemize}
-\item \verb \section   \item \verb \subsection  \item \verb \subsubsection  \item \verb \paragraph  \end{itemize}
-Section headings are automatically converted to allcaps style.
-\subsection{Lists}
-\label{sec:lists}
-%
-The \verb juliacon   class file provides unnumbered lists using the
-\verb unnumlist  environment for example,
-
-\begin{unnumlist}
-\item First unnumbered item which has no label and is indented from the
-left margin.
-\item Second unnumbered item.
-\item Third unnumbered item.
-\end{unnumlist}
-The unnumbered list which has no label and is indented from the
-left margin. was produced by:
-\begin{verbatim}
-\begin{unnumlist}
-\item First unnumbered item...
-\item Second unnumbered item...
-\item Third unnumbered item...
-\end{unnumlist}
-\end{verbatim}
-
-The \verb juliacon   class file also provides hyphen list using the
-\verb itemize  environment for example,
-\begin{itemize}
-\item First unnumbered bulleted item which has no label and is indented
-from the left margin.
-\item Second unnumbered bulleted item.
-\item Third unnumbered bulleted item which has no label and is indented
-from the left margin.
-\end{itemize}
-was produced by:
-\begin{verbatim}
-\begin{itemize}
-\item First item...
-\item Second item...
-\item Third item...
-\end{itemize}
-\end{verbatim}
-
-Numbered list is also provided in acmtog class file using the
-enumerate environment for example,
-\begin{enumerate}
-\item The attenuated and diluted stellar radiation.
-\item Scattered radiation, and
-\item Reradiation from other grains.
-\end{enumerate}
-
-was produced by:
-\begin{verbatim}
-\begin{enumerate}
-\item The attenuated...
-\item Scattered radiation, and...
-\item Reradiation from other grains...
-\end{enumerate}
-\end{verbatim}
-\subsection{Illustrations (or figures)}
-\label{subsub:sec_Illus}
-The \verb juliacon   class file will cope with most of the positioning of
-your illustrations and you should not normally use the optional positional
-qualifiers on the \verb figure   environment that would override
-these decisions.
-\vskip 6pt
-
-%
-\begin{figure}[t]
-\centerline{\includegraphics[width=4cm]{juliagraphs.png}}
-\caption{This is example of the image in a column.}
-	\label{fig:sample_figure}
-\end{figure}
-
-The figure \ref{fig:sample_figure} is taken from the JuliaGraphs
-organization \footnote{https://github.com/JuliaGraphs}.
-
-Figure captions should be \emph{below} the figure itself, therefore the
-\verb \caption  command should appear after the figure or space left for
-an illustration. For example, Figure 1 is produced using the following
-commands:
-
-\begin{verbatim}
-\begin{figure}
-\centerline{\includegraphics[width=20pc]{Graphics.eps}}
-\caption{An example of the testing process for a
-binary tree. The globa null hypothesis is tested
-first at level $\alpha$ (a), and the level of
-individual variables is reached last (d). Note
-that individual hypotheses can be tested at
-level $\alpha/4$ and not $\alpha/8$ as one might
-expect at first.}
-\label{sample-figure_2}
-\end{figure}
-\end{verbatim}
-Figures can be resized using first and second argument of
-\verb \includegraphics   command. First argument is used for modifying
-figure height and the second argument is used for modifying
-figure width respectively.
-\vskip 6pt
-Cross-referencing of figures, tables, and numbered, displayed
-equations using the \verb \label  and \verb \ref   commands is encouraged.
-For example, in referencing Figure 1 above, we used
-\verb Figure~\ref{sample-figure}   \subsection{Tables}
-\label{subsub:sec_Tab}
-The \verb juliacon   class file will cope with most of the positioning of
-your tables and you should not normally use the optional positional qualifiers on the table environment which would override these
-decisions. Table captions should be at the top.
-\begin{verbatim}
-\begin{table}
-\tbl{Tuning Set and Testing Set}{
-\begin{tabular}{|l|l|c|c|}\hline
-Label & \multicolumn{1}{c|}{Description}
-& Number of Users &
-Number of Queries\\\hline
-Train70 & Training Data &
-\smash{\raise-7pt\hbox{70}} & 104\\
-\cline{1-2}\cline{4-4}
-Test70 & Testing Data I & & 105\\\hline
-Test30 & Testing Data II & 30 & 119\\\hline
-& Total & 100 & 328\\\hline
-\end{tabular}}
-\end{table}\end{verbatim}
-
-\begin{table}
-\tbl{Tuning Set and Testing Set}{
-\begin{tabular}{|l|l|c|c|}\hline
-Label & \multicolumn{1}{c|}{Description}
-& Number of Users &
-Number of Queries\\\hline
-Test 1 & Training Data &
-\smash{\raise-7pt\hbox{70}} & 104\\
-\cline{1-2}\cline{4-4}
-Test 2 & Testing Data I & & 105\\\hline
-Test 3 & Testing Data II & 30 & 119\\\hline
-& Total & 100 & 328\\\hline
-\end{tabular}}
-\end{table}
-\subsection{Landscaping Pages}
-\label{subsub:landscaping_pages}
-If a table is too wide to fit the standard measure, it may be turned,
-with its caption, to 90 degrees. Landscape tables cannot be produced
-directly using the \verb juliacon   class file because \TeX{} itself cannot
-turn the page, and not all device drivers provide such a facility.
-The following procedure can be used to produce such pages.
-\vskip 6pt
-Use the package \verb rotating   in your document and change the coding
-from
-\begin{verbatim}
-\begin{table}...\end{table}
-to
-\begin{sidewaystable}...\end{sidewaystable}
-and for figures
-\begin{figure}...\end{figure}
-to
-\begin{sidewaysfigure}...\end{sidewaysfigure}
-\end{verbatim}
-
-environments in your document to turn your table on the appropriate
-page of your document. For instance, the following code prints
-a page with the running head, a message half way down and the
-table number towards the bottom.
-\begin{verbatim}
-\begin{sidewaystable}
-\tbl{Landscape table caption to go here.}{...}
-\label{landtab}
-\end{sidewaystable}
-\end{verbatim}
-
-\subsection{Double Column Figure and Tables}
-\label{subsub:double_fig_tab}
-For generating the output of figures and tables in double column
-we can use the following coding:
-
-\begin{enumerate}
-\item For Figures:
-\begin{verbatim}
-\begin{figure*}...\end{figure*}
-\end{verbatim}
-\item For landscape figures:
-\begin{verbatim}
-\begin{sidewaysfigure*}...\end{sidewaysfigure*}
-\end{verbatim}
-\item For Tables:
-\begin{verbatim}
-\begin{table*}...\end{table*}
-\end{verbatim}
-\item For landscape tables:
-\begin{verbatim}
-\begin{sidewaystable*}...\end{sidewaystable*}
-\end{verbatim}
-\end{enumerate}
-
-\subsection{Typesetting Mathematics}
-\label{subsub:type_math}
-The \verb juliacon class file will set displayed mathematics with center to
-the column width, provided that you use the \LaTeXe{} standard of
-open and closed square brackets as delimiters.
-The equation
-\[
-\sum_{i=1}^p \lambda_i = (S)
-\]
-
-was typeset using the acmtog class file with the commands
-
-\begin{verbatim}
-\[
-\sum_{i=1}^p \lambda_i = (S)
-\]
-\end{verbatim}
-
-For display equations, cross-referencing is encouraged. For example,
-\begin{verbatim}
-\begin{equation}
-(n-1)^{-1} \sum^n_{i=1} (X_i - \overline{X})^2.
-\label{eq:samplevar}
-\end{equation}
-Equation~(\ref{eq:samplevar}) gives the formula for
-sample variance.
-\end{verbatim}
-The following output is generated with the above coding:
-\begin{equation}
-(n-1)^{-1} \sum^n_{i=1} (X_i - \overline{X})^2.
-\label{eq:samplevar}
-\end{equation}
-Equation~(\ref{eq:samplevar}) gives the formula for
-sample variance.
-
-\subsection{Enunciations}
-\label{subsub:enunciation}
-The \verb juliacon   class file generates the enunciations with the help of
-the following commands:
-\begin{verbatim}
-\begin{theorem}...\end{theorem}
-\begin{strategy}...\end{strategy}
-\begin{property}...\end{property}
-\begin{proposition}...\end{proposition}
-\begin{lemma}...\end{lemma}
-\begin{example}...\end{example}
-\begin{proof}...\end{proof}
-\begin{definition}...\end{definition}
-\begin{algorithm}...\end{algorithm}
-\begin{remark}...\end{remark}
-\end{verbatim}
-The above-mentioned coding can also include optional arguments
-such as
-\begin{verbatim}
-\begin{theorem}[...]. Example for theorem:
-\begin{theorem}[Generalized Poincare Conjecture]
-Four score and seven ... created equal.
-\end{theorem}
-\end{verbatim}
-
-\begin{theorem}[Generalized Poincare Conjecture]
-Four score and seven years ago our fathers brought forth,
-upon this continent, a new nation, conceived in Liberty,
- and dedicated to the proposition that all men are
-created equal.
-\end{theorem}
-
-\subsection{Extract}
-\label{subsub:extract}
-Extract environment should be coded within
-\begin{verbatim}
-\begin{extract}..\end{extract}
-\end{verbatim}
-
-\subsection{Balancing column at last page}
-\label{subsub:Balance}
-For balancing the both column length at last page use :
-\begin{verbatim}
-\vadjust{\vfill\pagebreak}
-\end{verbatim}
-
-%\vadjust{\vfill\pagebreak}
-
-at appropriate place in your \TeX{} file or in bibliography file.
-
-\section{Handling references}
-\label{subsub:references}
-References are most easily (and correctly) generated using the
-BIBTEX, which is easily invoked via
-\begin{verbatim}
-\bibliographystyle{juliacon}
-\bibliography{ref}
-\end{verbatim}
-When submitting the document source (.tex) file to external
-parties, the ref.bib file should be sent with it.
-\cite{bezanson2017julia}
-
 \input{bib.tex}

### 2019-08-29 paper.tex
delete garbage
better cite
stuff

@@
 
-Using the Measurements.jl ~\cite{giordano2016} package to represent physical measurements with some uncertainty.
-JuliaDB is unaware of the Measurement datatype, yet it can store measurements, read and write to disk, and perform any database operation while using the math defined on the measurement type.
-In a two-language system, a performant implementation of the measurement datastructure must be done in a low-level language like C.
-For performant queries on columns containing measurements, one will have to implement an array of measurements datastructure in the low-level language as well that supports many array operations.
-But with JuliaDB and Julia, simply defining the measurement types and the primitive operations on Measurements is enough to achieve all the said features.
-Measurements can be stored efficiently in Julia arrays, operations such as `mean` and `sum` or broadcast on arrays of measurements fall out their existing generic implementations due to Julia's dynamic multiple-dispatch based JIT compiler.
+We take up a prototypical illustration of the power of JuliaDB by using a custom number type, namely the Measurement type, in a database.
 
-Measurements.jl implements the ± operator
-which represents an observation with some uncertainty associated
-with it. Numeric operations on measuements propagate the
-uncertainty as prescribed by linear error propagation theory:
-For a function application $G(a,b,c...)$ where arguments are measurements $a=\overline{a} \pm \sigma_{a}, b=\overline{b} \pm \sigma_{b}, c=\overline{c} \pm \sigma_{c}$ the error of the output is found using the relationship:
+Measurements.jl implements the ± operator which represents an observation with some uncertainty associated with it.
+Numeric operations on measuements propagate the uncertainty as prescribed by linear error propagation theory: For a function application $G(a,b,c...)$ where arguments are measurements $a=\overline{a} \pm \sigma_{a}, b=\overline{b} \pm \sigma_{b}, c=\overline{c} \pm \sigma_{c}$ the error of the output is found using the relationship:
 
-$$
-\begin{aligned}\sigma_{G}^{2}=&\left(\left.\frac{\partial G}{\partial a}\right|_{a=\overline{a}} \sigma_{a}\right)^{2}+\left(\left.\frac{\partial G}{\partial b}\right|_{b=\overline{b}} \sigma_{b}\right)^{2}+\left(\left.\frac{\partial G}{\partial c}\right|_{c=\overline{c}} \sigma_{c}\right)^{2}+\cdots \\
-&+2\left(\frac{\partial G}{\partial a}\right)_{a=\overline{a}}\left(\frac{\partial G}{\partial b}\right)_{b=\overline{b}} \sigma_{a b}+2\left(\frac{\partial G}{\partial a}\right)_{a=\overline{a}}\left(\frac{\partial G}{\partial c}\right)_{c=\overline{c}} \sigma_{a c} \\ &+2\left(\frac{\partial G}{\partial b}\right)_{b=\overline{b}}\left(\frac{\partial G}{\partial c}\right)_{c=\overline{c}} \sigma_{b c}+\cdots \end{aligned}
-$$\\
+\begin{equation}
+$$\sigma_{G}^{2}=\left(\left.\frac{\partial G}{\partial a}\right|_{a=\overline{a}} \sigma_{a}\right)^{2}+\left(\left.\frac{\partial G}{\partial b}\right|_{b=\overline{b}} \sigma_{b}\right)^{2}+\left(\left.\frac{\partial G}{\partial c}\right|_{c=\overline{c}} \sigma_{c}\right)^{2}+\cdots \\
++2\left(\frac{\partial G}{\partial a}\right)_{a=\overline{a}}\left(\frac{\partial G}{\partial b}\right)_{b=\overline{b}} \sigma_{a b}+2\left(\frac{\partial G}{\partial a}\right)_{a=\overline{a}}\left(\frac{\partial G}{\partial c}\right)_{c=\overline{c}} \sigma_{a c} \\ +2\left(\frac{\partial G}{\partial b}\right)_{b=\overline{b}}\left(\frac{\partial G}{\partial c}\right)_{c=\overline{c}} \sigma_{b c}+\cdots$$
+
+\end{equation}
 
@@
 allocation and parallelism.
+
+The Measurement type in the Measurements.jl ~\cite{giordano2016} package represents physical measurements with some uncertainty.
+JuliaDB is unaware of the Measurement datatype, yet it can store, read and write to disk, and perform any database operation while using the math defined on the measurement type.
+In a two-language system, a performant implementation of the measurement datastructure must be done in a low-level language like C.
+For reasonable performance on queries on columns containing measurements, one may have to implement an array of measurements datastructure in the low-level language as well that supports many array operations.
+But with JuliaDB and Julia, simply defining the measurement type and the primitive operations on Measurements is enough to achieve all the said features.
+Measurements can be stored efficiently in Julia arrays, operations such as `mean` and `sum` or broadcast on arrays of measurements fall out their existing generic implementations due to Julia's dynamic multiple-dispatch based JIT compiler.
 

### 2019-08-29 paper.tex
add Measurements demo listing

@@
 
-\begin{equation}
-$$\sigma_{G}^{2}=\left(\left.\frac{\partial G}{\partial a}\right|_{a=\overline{a}} \sigma_{a}\right)^{2}+\left(\left.\frac{\partial G}{\partial b}\right|_{b=\overline{b}} \sigma_{b}\right)^{2}+\left(\left.\frac{\partial G}{\partial c}\right|_{c=\overline{c}} \sigma_{c}\right)^{2}+\cdots \\
-+2\left(\frac{\partial G}{\partial a}\right)_{a=\overline{a}}\left(\frac{\partial G}{\partial b}\right)_{b=\overline{b}} \sigma_{a b}+2\left(\frac{\partial G}{\partial a}\right)_{a=\overline{a}}\left(\frac{\partial G}{\partial c}\right)_{c=\overline{c}} \sigma_{a c} \\ +2\left(\frac{\partial G}{\partial b}\right)_{b=\overline{b}}\left(\frac{\partial G}{\partial c}\right)_{c=\overline{c}} \sigma_{b c}+\cdots$$
-
+$$\begin{equation}
+  \begin{split}
+    \sigma_G^2 &= \left( \left.\frac{\partial G}{\partial a}\right\vert_{a =
+        \bar{a}} \sigma_a \right)^2 + \left( \left.\frac{\partial G}{\partial
+          b}\right\vert_{b = \bar{b}} \sigma_b \right)^2 + \left(
+      \left.\frac{\partial G}{\partial c}\right\vert_{c = \bar{c}} \sigma_c
+    \right)^2 + \cdots \\
+    &\quad{}+ 2 \left(\frac{\partial G}{\partial a}\right)_{a = \bar{a}}
+    \left(\frac{\partial G}{\partial b}\right)_{b = \bar{b}} \sigma_{ab} + 2
+    \left(\frac{\partial G}{\partial a}\right)_{a = \bar{a}}
+    \left(\frac{\partial G}{\partial c}\right)_{c = \bar{c}}
+    \sigma_{ac} \\
+    &\quad{}+ 2 \left(\frac{\partial G}{\partial b}\right)_{b = \bar{b}}
+    \left(\frac{\partial G}{\partial c}\right)_{c = \bar{c}} \sigma_{bc} +
+    \cdots
+  \end{split}
 \end{equation}
+$$
 
@@
 
-Unlike a typical non-Julia analytical databases, without any
+Unlike a typical non-Julia analytical database, without any
 first class support for Measurements,
@@
 
+\begin{lstlisting}[language=Julia]
+# generate some fake measurements time-series:
+using JuliaDB, Measurements, Dates, Random
+Random.seed!(1234)
+
+n = 100000
+timestamps = sort(DateTime.(now() .- Millisecond.(rand(1:10^6, n))))
+x = cumsum(randn(n)) .± rand(n)./10
+
+#=
+Table with 100000 rows, 2 columns:
+time                     x
+────────────────────────────────────────
+2019-08-29T01:52:44.148  -1.222±0.086
+2019-08-29T01:52:44.178  -0.007±0.02
+2019-08-29T01:52:44.201  0.801±0.053
+...
+=#
+\end{lstlisting}
+
 Perform groupby-mean without implementing the
@@
 
-Use OnlineStats for stats on measurements for lesser
-allocation and parallelism.
+\begin{lstlisting}[language=Julia]
+minutely = groupby(mean, observations,
+                  :time => t->trunc(t, Minute), select=:x)
+
+#=
+Table with 18 rows, 2 columns:
+time                 mean
+─────────────────────────────────────
+2019-08-29T01:52:00  28.1203±0.0015
+2019-08-29T01:53:00  59.9434±0.00075
+2019-08-29T01:54:00  4.21493±0.00075
+...
+=#
+\end{lstlisting}
+
+Use OnlineStats for stats on measurements for lesser allocation and parallelism.
+
+\begin{lstlisting}
+# assuming session was started with many julia workers
+
+using OnlineStats
+
+d_observations = distribute(observations, 8) # distribute into 8 parts, equally split between available julia processes
+d_minutely = groupreduce(Mean(Measurement{Float64}), d_observations,
+                         :time => t->trunc(t, Minute), select=:x)
+
+#=
+Distributed Table with 18 rows in 1 chunks:
+time                 mean
+─────────────────────────────────────
+2019-08-29T01:52:00  Mean: n=1577 | value=28.1203±0.0015
+2019-08-29T01:53:00  Mean: n=5989 | value=59.9434±0.00075
+2019-08-29T01:54:00  Mean: n=5883 | value=4.21493±0.00075
+...
+=#
+\end{lstlisting}
 

### 2019-08-29 paper.tex
"generate" header.tex by hand

@@
 \begin{lstlisting}
-# assuming session was started with many julia workers
+# assuming Julia session was started with many julia workers
 
@@
 
-d_observations = distribute(observations, 8) # distribute into 8 parts, equally split between available julia processes
-d_minutely = groupreduce(Mean(Measurement{Float64}), d_observations,
-                         :time => t->trunc(t, Minute), select=:x)
+# distribute into 8 parts, equally split
+# between available julia processes
+d_observations = distribute(observations, 8)
+d_minutely = groupreduce(Mean(Measurement{Float64}),
+                         d_observations,
+                         :time => t->trunc(t, Minute),
+                         select=:x)
 

### 2019-08-29 paper.tex
lstset
things

@@
 
-n = 100000
+function random_observations(n)
     timestamps = sort(DateTime.(now() .- Millisecond.(rand(1:10^6, n))))
     x = cumsum(randn(n)) .± rand(n)./10
+    table((time=timestamps, x=x))
+end
+
+observations = random_observations(100000)
 
@@
 \end{lstlisting}
+
+Finally, the Measurements.jl defines how plotting must be performed for an array of measurements, hence, for example a plotting tool that knows the tabular structure of JuliaDB, such as StatPlots.jl can be readily used to plot the table.
+The plotting automatically shows error bars representing uncertainty.
 
@@
 
+\section{Benchmarks}
+
 Here we show benchmarks of JuliaDB against other popular data science

### 2019-08-29 paper.tex
start application section
alan

@@
 \setcounter{page}{1}
+
+\lstset{literate=
+{─}{{\ucc{$--$}}}{1}
+{±}{{\ucc{$\pm$}}}{1}
+}
 
@@
 the database implementer greatly.
+
+\section{Design of JuliaDB}
+
+In the following subsection we describe the design of JuliaDB which empowers performance and interoperability.
+
+\subsection{Columnar memory layout}
+The basic datastructure in JuliaDB is the table.
+
+\begin{lstlisting}[language=Julia]
+lotr = table((name=StringArray(["Frodo", "Samwise", "Bilbo",
+                                "Aragorn", "Boromir", "Gimli",
+                                "Elrond", "Legolas"]),
+             race=PooledArray(["Hobbit", "Hobbit", "Hobbit",
+                               "Human", "Human", "Dwarf",
+                               "Elf", "Elf"]),
+             age=[33, 61, 111, 87, 41, 139, 6000, 2931]));
+
+collect(lotr)
+
+#=
+8-element Array{NamedTuple{(:name, :race, :age),
+                Tuple{String,String,Int64}},1}:
+ (name = "Frodo", race = "Hobbit", age = 33)  
+ (name = "Samwise", race = "Hobbit", age = 61)
+ (name = "Bilbo", race = "Hobbit", age = 111) 
+ (name = "Aragorn", race = "Human", age = 87) 
+ (name = "Boromir", race = "Human", age = 41) 
+ (name = "Gimli", race = "Dwarf", age = 139)  
+ (name = "Elrond", race = "Elf", age = 6000)  
+ (name = "Legolas", race = "Elf", age = 2931) 
+=#
+\end{lstlisting}
+Listing 1. \emph{A table stores a named tuple of columns, but produces named tuple rows when iterated.}\\
+
+Listing 1 shows a table constructed from a named tuple of column vectors.
+Although a table stores data as a named tuple of columns, it acts as an iterator of rows.
+Rows are tuples or named tuples of items.
+
+The columns can be accessed using the \texttt{columns} function (e.g.
+\texttt{rows(lotr)} returns a named tuple of column vectors provided to \texttt{table} during construction) while the rows can be accessed as an array of tuples using the \texttt{rows} function rows is a struct of arrays view into the columnar storage, unlike \texttt{table} which returns a JuliaDB table, \texttt{rows} returns a legitimate Julia array.
+
+Tables support \texttt{map}, \texttt{reduce}, \texttt{groupreduce}, \texttt{groupby} and \texttt{join} operations.
+
+Listing 1 also illustrates use of \texttt{StringArray} and \texttt{PooledArray} array types as columns. These special array types enable efficient packing of String and categorical data respectively.
+
+\begin{figure*}[t]
+\centerline{\includegraphics[width=11cm]{stringarray.png}}
+\caption{StringArray layout.}
+  \label{fig:stringarray}
+\end{figure*}
+
+A \texttt{StringArray}'s memory layout is show in in Figure \ref{fig:stringarray}.
+The contents of strings (unicode character codes) are packed into a buffer of 8-bit unsigend integers.
+\texttt{StringArray} also stores the lengths of each string and the index (offset) at which each of them starts.
+Although it would suffice to store just the length of each string, storing offset also allows us to implement efficient mutation operations.
+Setting an element of the array appends the new string to the end of the buffer and replaces the length and offset with the corresponding new length and offsets for that item.
+Operations such as \texttt{sort!} and \texttt{permute!} (in place permutation of items) can simply modify the length and offset arrays and leave the buffer unmodified.
+Wherever suitable, in code internal to JuliaDB, we create \texttt{WeakRefString} instead of creating String objects when we index into a \texttt{StringArray}, this greatly reduces GC-load when dealing with string columns in operations such as groupby and join.
+
+A \texttt{PooledArray}'s memory layout is shown in Figure \ref{fig:pooledarray}.
+It is useful in compressing data with low number of unique items as compared to the total number of items.
+PooledArrays stores a dictionary mapping each unique item to a \texttt{ref} integer, and then an array of refs which tells which unique element appears in what position in the uncompressed array.
+
+\subsection{Sort-based indexing}
+
+We implement groupby and join by traversing tables in the key-order.
+We use counting sort wherever possible (e.g.
+integer columns, PooledArrays) to do a small amount of initial work on the order of $O(k \log k)$ where $k$ is the number of unique values and then acheive a sorted permutation in $O n$ time where $n$ is the number of rows.
+In most systems where the implementation is done in C++, the ordering routines are restricted to a handful of primitive data types.
+In our system any user defined data type with an \code{isless} method to compare two items can take part in the sorting.
+
+\textbf{Indexing: }For speeding up table operations, you can tell
+JuliaDB to keep the rows of a table sorted by some of its fields.
+These fields are called primary key fields. The primary key fields
+can be of any data type which has an \code{isless} method associated
+with it to compare two values. This includes but is not limited to
+numbers, strings, symbols, dates, time periods, currencies, quantities
+of comparable physical units (e.g. inch, mm).
+
+\subsection{Distributed compute}
+
+A distributed table is a distributed analogue of the one described
+above. It is stored as many tables across multiple Julia processes,
+\code{columns} returns a tuple of \emph{distributed arrays} and
+\code{rows} returns a distributed array of tuples! The distributed
+arrays implementation has the usual array operations (map, reduce,
+hcat, vcat) defined on it in a distributed manner. Constructing a
+table with one or more distributed columns using the table constructor
+results in a distributed table. We use the word ``table'' from now on to refer interchangeably to both types of tables, unless
+noted otherwise.
 
@@
 
-\section{Design of JuliaDB}
-
-In the following subsection we describe the design of JuliaDB which empowers performance and interoperability.
-
-\subsection{Columnar memory layout}
-The basic datastructure in JuliaDB is the table.
-
-\begin{lstlisting}[language=Julia]
-lotr = table((name=StringArray(["Frodo", "Samwise", "Bilbo",
-                                "Aragorn", "Boromir", "Gimli",
-                                "Elrond", "Legolas"]),
-             race=PooledArray(["Hobbit", "Hobbit", "Hobbit",
-                               "Human", "Human", "Dwarf",
-                               "Elf", "Elf"]),
-             age=[33, 61, 111, 87, 41, 139, 6000, 2931]));
-
-collect(lotr)
-
-#=
-8-element Array{NamedTuple{(:name, :race, :age),
-                Tuple{String,String,Int64}},1}:
- (name = "Frodo", race = "Hobbit", age = 33)  
- (name = "Samwise", race = "Hobbit", age = 61)
- (name = "Bilbo", race = "Hobbit", age = 111) 
- (name = "Aragorn", race = "Human", age = 87) 
- (name = "Boromir", race = "Human", age = 41) 
- (name = "Gimli", race = "Dwarf", age = 139)  
- (name = "Elrond", race = "Elf", age = 6000)  
- (name = "Legolas", race = "Elf", age = 2931) 
-=#
-\end{lstlisting}
-Listing 1. \emph{A table stores a named tuple of columns, but produces named tuple rows when iterated.}\\
-
-Listing 1 shows a table constructed from a named tuple of column vectors.
-Although a table stores data as a named tuple of columns, it acts as an iterator of rows.
-Rows are tuples or named tuples of items.
-
-The columns can be accessed using the \texttt{columns} function (e.g.
-\texttt{rows(lotr)} returns a named tuple of column vectors provided to \texttt{table} during construction) while the rows can be accessed as an array of tuples using the \texttt{rows} function rows is a struct of arrays view into the columnar storage, unlike \texttt{table} which returns a JuliaDB table, \texttt{rows} returns a legitimate Julia array.
-
-Tables support \texttt{map}, \texttt{reduce}, \texttt{groupreduce}, \texttt{groupby} and \texttt{join} operations.
-
-Listing 1 also illustrates use of \texttt{StringArray} and \texttt{PooledArray} array types as columns. These special array types enable efficient packing of String and categorical data respectively.
-
-\begin{figure*}[t]
-\centerline{\includegraphics[width=11cm]{stringarray.png}}
-\caption{StringArray layout.}
-  \label{fig:stringarray}
-\end{figure*}
-
-A \texttt{StringArray}'s memory layout is show in in Figure \ref{fig:stringarray}.
-The contents of strings (unicode character codes) are packed into a buffer of 8-bit unsigend integers.
-\texttt{StringArray} also stores the lengths of each string and the index (offset) at which each of them starts.
-Although it would suffice to store just the length of each string, storing offset also allows us to implement efficient mutation operations.
-Setting an element of the array appends the new string to the end of the buffer and replaces the length and offset with the corresponding new length and offsets for that item.
-Operations such as \texttt{sort!} and \texttt{permute!} (in place permutation of items) can simply modify the length and offset arrays and leave the buffer unmodified.
-Wherever suitable, in code internal to JuliaDB, we create \texttt{WeakRefString} instead of creating String objects when we index into a \texttt{StringArray}, this greatly reduces GC-load when dealing with string columns in operations such as groupby and join.
-
-A \texttt{PooledArray}'s memory layout is shown in Figure \ref{fig:pooledarray}.
-It is useful in compressing data with low number of unique items as compared to the total number of items.
-PooledArrays stores a dictionary mapping each unique item to a \texttt{ref} integer, and then an array of refs which tells which unique element appears in what position in the uncompressed array.
-
-\subsection{Sort-based indexing}
-
-We implement groupby and join by traversing tables in the key-order.
-We use counting sort wherever possible (e.g.
-integer columns, PooledArrays) to do a small amount of initial work on the order of $O(k \log k)$ where $k$ is the number of unique values and then acheive a sorted permutation in $O n$ time where $n$ is the number of rows.
-In most systems where the implementation is done in C++, the ordering routines are restricted to a handful of primitive data types.
-In our system any user defined data type with an \code{isless} method to compare two items can take part in the sorting.
-
-\textbf{Indexing: }For speeding up table operations, you can tell
-JuliaDB to keep the rows of a table sorted by some of its fields.
-These fields are called primary key fields. The primary key fields
-can be of any data type which has an \code{isless} method associated
-with it to compare two values. This includes but is not limited to
-numbers, strings, symbols, dates, time periods, currencies, quantities
-of comparable physical units (e.g. inch, mm).
-
-\subsection{Distributed compute}
-
-A distributed table is a distributed analogue of the one described
-above. It is stored as many tables across multiple Julia processes,
-\code{columns} returns a tuple of \emph{distributed arrays} and
-\code{rows} returns a distributed array of tuples! The distributed
-arrays implementation has the usual array operations (map, reduce,
-hcat, vcat) defined on it in a distributed manner. Constructing a
-table with one or more distributed columns using the table constructor
-results in a distributed table. We use the word ``table'' from now on to refer interchangeably to both types of tables, unless
-noted otherwise.
-
 \section{Benchmarks}
@@
 
-\begin{figure}[h]
+\begin{figure*}[h]
 \centering \includegraphics[width=5in]{image8.png} \caption{Performance comparison for loading 815MB from the NYC taxi dataset}
 \label{fig:nyctaxiload} 
-\end{figure}
+\end{figure*}
 Our test for loading data is an 815 MB CSV file from the NYC Taxi
@@
 
-\begin{figure}[h]
+\begin{figure*}[h]
 \centering \includegraphics[width=5in]{image4.png} \caption{Query times for 3 different query types across various systems}
 \label{fig:nyctaxiquery} 
-\end{figure}
+\end{figure*}
+
+\section{Applications}
+
+We breifly mention some applications of JuliaDB in the real-world. Julia Computing is using JuliaDB as a time-series database~\cite{wilmott2018} for large datasets.
 

### 2019-08-29 paper.tex
abs

@@
 \begin{abstract}
-
-We present JuliaDB, a distributed database for analytical workloads that is built solely in Julia, a high-level high-performance dynamic language for numerical computing.
+We describe how the two-language problem manifests in the world of analytical databases.
+A system with this problem takes at least two languages to implement: a low-level statically compiled language such as C or C++ to implement the performance-sensitive operations, and a high-level dynamic scripting language such as Python or R to implement the user-facing API.
+This dichotomy usually exists because the latter language cannot compile to efficient machine code like the former, and the former does not have the ease-of-use as the latter.
+We present JuliaDB, a distributed database for analytical workloads that is built solely in Julia, and does not have this and other resulting problems.
 JuliaDB is designed to allow seamless use of user-defined Julia functions and data types.
 Julia's just-in-time compilation welds code from across Julia's package ecosystem as well as user code to generate efficient machine code allowing for high-performance data science in a coherent environment.
-We show example applications involving error analysis, and online statistics.
-Our benchmark shows that the system can perform on-par with or betters that of pandas and tidyverse in R.
+We show a case study involving the use of a custom numeric type.
+Our benchmark shows that the system can perform on-par with or betters that of pandas and tidyverse in the cases we tested.
+
 \headingtable

### 2019-08-29 paper.tex
Updates from Overleaf
captions and cite users

@@
+\usepackage{url}
 % JuliaCon proceedings template
@@
 This dichotomy usually exists because the latter language cannot compile to efficient machine code like the former, and the former does not have the ease-of-use as the latter.
-We present JuliaDB, a distributed database for analytical workloads that is built solely in Julia, and does not have this and other resulting problems.
+We present JuliaDB, a distributed database for analytical workloads that is built solely in Julia and does not have this and other resulting problems.
 JuliaDB is designed to allow seamless use of user-defined Julia functions and data types.
@@
     \label{fig:nyctaxiquery} 
+    \caption{Query performance comparison with other systems.
+Query 1: average fare amount by type of vendor.
+Query 2: number of trips grouped by number of passengers and day of week.
+Query 3: number of trips grouped by number of passengers and even and odd days of week.
+(Monday, Wednesday, Firday are even, rest are odd).
 \end{figure*}
 
-\section{Applications}
+\section{Usage in real world}
 
-We breifly mention some applications of JuliaDB in the real-world. Julia Computing is using JuliaDB as a time-series database~\cite{wilmott2018} for large datasets.
+Josh Day describes how JuliaDB is a natural time-series database~\cite{wilmott2018} for large datasets without having been designed with the specific purpose of time-series in mind.
+JuliaDB is used by the package Diversity.jl~\cite{claireh2018} to generate results in understanding the evolutionary relationships between species across the whole kingdom of flowering plants, and simulations to predict responses of plants species to climate change.
 

### 2019-08-29 paper.tex
code -> texttt

@@
-\usepackage{url}
 % JuliaCon proceedings template
@@
 We use counting sort wherever possible (e.g.
-integer columns, PooledArrays) to do a small amount of initial work on the order of $O(k \log k)$ where $k$ is the number of unique values and then acheive a sorted permutation in $O n$ time where $n$ is the number of rows.
+integer columns, PooledArrays) to do a small amount of initial work on the order of $O(k \log k)$ where $k$ is the number of unique values and then achieve a sorted permutation in $O n$ time where $n$ is the number of rows.
 In most systems where the implementation is done in C++, the ordering routines are restricted to a handful of primitive data types.
-In our system any user defined data type with an \code{isless} method to compare two items can take part in the sorting.
+In our system any user defined data type with an \texttt{isless} method to compare two items can take part in the sorting.
 
@@
 These fields are called primary key fields. The primary key fields
-can be of any data type which has an \code{isless} method associated
+can be of any data type which has an \texttt{isless} method associated
 with it to compare two values. This includes but is not limited to

### 2019-08-29 paper.tex
textttt
10 things

@@
 
+Storing data row-by-row is not ideal when a query needs to operate on a small subset of fields for in such cases, one would have to load the whole dataset into memory and wade through it to find the relavent fields.
+But if we stored the columns separately, a single field's data is all available together and hence queries requiring data from a subset of fields load only data for those fields.
+The only advantage of row-by-row storage would be to read multiple fields per row on a sequential access optimized device such as tape drive or a spinning-disk hard drive.
+It's not worth worrying about these cases anymore.
+
+The reason a table behaves as an iterator is a matter of convension.
+It's more helpful to think of writing code one row at a time.
+JuliaDB creates an immutable struct every time a row needs to be materialized.
+This may or may not lead to heap-allocation depending on the compiler optimizations that occur.
+In most cases the compiler is smart enough to eleminate allocation.
+
 The columns can be accessed using the \texttt{columns} function (e.g.
@@
 
-Tables support \texttt{map}, \texttt{reduce}, \texttt{groupreduce}, \texttt{groupby} and \texttt{join} operations.
+Tables naturally support \texttt{map}, \texttt{reduce}, \texttt{groupreduce}, \texttt{groupby} and \texttt{join} operations.
+We will defer their discussion to the API documentation of JuliaDB.
+Read the Sectoin \ref{sec:sort} for more about the implementation of some of these operations.
 
@@
 \subsection{Sort-based indexing}
+\label{sec:sort}
 
@@
 above. It is stored as many tables across multiple Julia processes,
-\code{columns} returns a tuple of \emph{distributed arrays} and
-\code{rows} returns a distributed array of tuples! The distributed
+\texttt{columns} returns a tuple of \emph{distributed arrays} and
+\texttt{rows} returns a distributed array of tuples! The distributed
 arrays implementation has the usual array operations (map, reduce,

### 2019-08-29 paper.tex
Updates from Overleaf

--- 2019-08-29 paper.tex
@@
 $$\begin{equation}
-  \begin{split}
+  \begin{align}
     \sigma_G^2 &= \left( \left.\frac{\partial G}{\partial a}\right\vert_{a =
@@
     \cdots
-  \end{split}
-\end{equation}
-$$
+  \end{align}
+\end{equation}$$
 
@@
 \begin{figure*}[h]
-    \centering \includegraphics[width=5in]{image4.png} \caption{Query times for 3 different query types across various systems}
+    \centering \includegraphics[width=5in]{image4.png}
     \label{fig:nyctaxiquery} 
@@
 Query 3: number of trips grouped by number of passengers and even and odd days of week.
-(Monday, Wednesday, Firday are even, rest are odd).
+(Monday, Wednesday, Firday are even, rest are odd).}
 \end{figure*}

### 2019-08-30 paper.tex
lots of stuff

--- 2019-08-29 paper.tex
@@
 We describe how the two-language problem manifests in the world of analytical databases.
-A system with this problem takes at least two languages to implement: a low-level statically compiled language such as C or C++ to implement the performance-sensitive operations, and a high-level dynamic scripting language such as Python or R to implement the user-facing API.
+A two-language system is one that is implemented using at least two languages: a low-level statically compiled language such as C or C++ to implement the performance-sensitive operations, and a high-level dynamic scripting language such as Python or R to implement the user-facing API.
 This dichotomy usually exists because the latter language cannot compile to efficient machine code like the former, and the former does not have the ease-of-use as the latter.
-We present JuliaDB, a distributed database for analytical workloads that is built solely in Julia and does not have this and other resulting problems.
+We present JuliaDB, a distributed analytical database written entirely in Julia.
+Julia removes the shortcomings of the said two classes of languages: it is both dynamic and easy to use, and compiles to efficient machine code.
 JuliaDB is designed to allow seamless use of user-defined Julia functions and data types.
-Julia's just-in-time compilation welds code from across Julia's package ecosystem as well as user code to generate efficient machine code allowing for high-performance data science in a coherent environment.
-We show a case study involving the use of a custom numeric type.
-Our benchmark shows that the system can perform on-par with or betters that of pandas and tidyverse in the cases we tested.
+Julia's just-in-time compilation welds code from across Julia's package ecosystem as well as user code to generate efficient machine code allowing for systems that are greater than the sum of parts.
+We show a case study involving the use of measurement type to propagate uncertainty in measurements.
+Our benchmark shows that the system can perform on-par with or better than two-language systems like pandas and tidyverse in the cases we tested.
 
@@
 and efficient manually-memory-managed languages such as C or C++ helping
-the database implementer greatly.
+the database implementer and user greatly.
 
@@
 In the following subsection we describe the design of JuliaDB which empowers performance and interoperability.
+It is a testament to Julia's power as an implementation of JuliaDB that we were able to address all problems except "eager evaluation" from the list of problems originally thought to have been seen, some from omission and some due to the two-language divide \cite{wes2017}.
+We leave the implementation of lazy-evaluated queries for future work.
 
@@
 
+Storing data row-by-row is not ideal when a query needs to operate on a small subset of fields, for in such cases, one would have to load the whole dataset into memory and wade through it to find the relavent fields.
+But if we stored the columns separately, a single field's data is all available together, and in simple cases, contiguously in memory.
+Hence queries requiring data from a subset of fields load only data for those columns.
+The only advantage of row-by-row storage would be to read multiple fields per row on a sequential access storage device such as tape drive or a spinning-disk hard drive.
+It's not worth worrying about these cases for an analytical database mostly interested in performance.
+
+When a table is iterated, it yields named tuples (rows).
+The reason a table behaves as an iterator of named tuples is a matter of convension.
+JuliaDB creates an immutable struct every time a row needs to be materialized.
+This may or may not lead to heap-allocation depending on the compiler optimizations that become possible.
+
+The type of the row tuple is encoded in that of the table itself, hence the Julia compiler always has information about what type a certain field of a certain table is, and is able to propagate it to produce efficient machine code.
+
 The columns can be accessed using the \texttt{columns} function (e.g.
@@
 
-Tables support \texttt{map}, \texttt{reduce}, \texttt{groupreduce}, \texttt{groupby} and \texttt{join} operations.
+Naturally, tables support \texttt{map}, \texttt{reduce}, \texttt{groupreduce}, \texttt{groupby} and \texttt{join} operations.
+We will defer their discussion to the API documentation of JuliaDB \cite{juliadbapi}.
+Read the Sectoin \ref{sec:sort} for more about the implementation of some of these operations.
 
@@
 
-A \texttt{StringArray}'s memory layout is show in in Figure \ref{fig:stringarray}.
-The contents of strings (unicode character codes) are packed into a buffer of 8-bit unsigend integers.
+A \texttt{StringArray}'s stores a collection of strings with less memory indirection than a usual array of strings.
+Its memory layout is show in in Figure \ref{fig:stringarray}.
+The contents of strings (unicode character codes) are packed into a contiguous buffer of 8-bit unsigend integers.
 \texttt{StringArray} also stores the lengths of each string and the index (offset) at which each of them starts.
@@
 Wherever suitable, in code internal to JuliaDB, we create \texttt{WeakRefString} instead of creating String objects when we index into a \texttt{StringArray}, this greatly reduces GC-load when dealing with string columns in operations such as groupby and join.
+Some advantages of a StringArray are 1) the buffer, and the length and offset vectors can be memory mapped to disk.
+2) it takes less serialization overhead to send the same over the wire to another process.
 
-A \texttt{PooledArray}'s memory layout is shown in Figure \ref{fig:pooledarray}.
-It is useful in compressing data with low number of unique items as compared to the total number of items.
-PooledArrays stores a dictionary mapping each unique item to a \texttt{ref} integer, and then an array of refs which tells which unique element appears in what position in the uncompressed array.
+If the number of unique elements in an array is much less than the number of elements in an array, it is benefitial to store it as a PooledArray.
+A PooledArray stores 1) a dictionary mapping each unique element to an integer (the ref of that value) 2) a dictionary mapping ref to its value (the inverse of 1)) and 3) an array of refs which denotes which unique value appears at what position in the actual data.
+This is a good compression mechanism for the data as well.
+PooledArrays also have efficient sorting algorithms as described in the next sub section.
+PooledArray refs can be memory mapped to and from disk.
 
 \subsection{Sort-based indexing}
+\label{sec:sort}
 
-We implement groupby and join by traversing tables in the key-order.
+We implement \texttt{groupby} and \texttt{join} by traversing tables in the key-order.
+For example, a groupby operation first finds a permutation of the table which will result in the groupby key fields being sorted, and traveses the table finding contiguous blocks of equal keys, and performs the grouping operation on each contiguous block.
+The join operation works by traversing two tables in lock-step and combining relavent rows to the type of join being performed and generating the output table.
+
 We use counting sort wherever possible (e.g.
-integer columns, PooledArrays) to do a small amount of initial work on the order of $O(k \log k)$ where $k$ is the number of unique values and then achieve a sorted permutation in $O n$ time where $n$ is the number of rows.
+integer columns that have acceptable extrema).
+On PooledArrays, we do a small amount of initial work in the order of $O(k \log k)$ where $k$ is the number of unique values and then find the sorted permutation in $O(n)$ time where $n$ is the number of rows.
 In most systems where the implementation is done in C++, the ordering routines are restricted to a handful of primitive data types.
@@
 
-\subsection{Distributed compute}
+\subsection{Parallelism}
 
 A distributed table is a distributed analogue of the one described
-above. It is stored as many tables across multiple Julia processes,
-\code{columns} returns a tuple of \emph{distributed arrays} and
-\code{rows} returns a distributed array of tuples! The distributed
-arrays implementation has the usual array operations (map, reduce,
-hcat, vcat) defined on it in a distributed manner. Constructing a
+above. It is stored as many tables across multiple Julia processes. On distributed tables,
+\texttt{columns} returns a tuple of \emph{distributed arrays} and
+\texttt{rows} returns a distributed array of tuples! The distributed
+arrays implementation has the usual array operations (\texttt{map}, \texttt{reduce}, \texttt{hcat}, \texttt{vcat}) defined on it. Constructing a
 table with one or more distributed columns using the table constructor
@@
 
-We take up a prototypical illustration of the power of JuliaDB by using a custom number type, namely the Measurement type, in a database.
+We take up a prototypical illustration of the power of JuliaDB by creating a table with data of a custom numeric type, namely the \texttt{Measurement} type provided by the Measurments.jl~\cite{giordano2016} package.
 
-Measurements.jl implements the ± operator which represents an observation with some uncertainty associated with it.
+Measurements.jl implements the ± operator which takes two numbers -- the mean of the observation and the uncertainty, and constructs a \texttt{Measurement} object.
 Numeric operations on measuements propagate the uncertainty as prescribed by linear error propagation theory: For a function application $G(a,b,c...)$ where arguments are measurements $a=\overline{a} \pm \sigma_{a}, b=\overline{b} \pm \sigma_{b}, c=\overline{c} \pm \sigma_{c}$ the error of the output is found using the relationship:
@@
 
-Unlike a typical non-Julia analytical database, without any
-first class support for Measurements,
-Pure-julia analytical database
-We can store measurements directly in a JuliaDB table.
+Unlike in two-language analytical systems, JuliaDB need not be aware of the Measurements package as long as it implements a legitimate number type.
+Many features become available the moment JuliaDB and Measurements packages are included into an environment:
+
+\begin{enumerate}
+    \item We can store measurements directly in a JuliaDB table.
 
@@
 
-Perform groupby-mean without implementing the
+\item Perform groupby-mean without implementing the
 mean function for an array of measurements in C.
@@
 
-Use OnlineStats for stats on measurements for lesser allocation and parallelism.
+\item Use OnlineStats for stats on measurements for lesser allocation and parallelism.
 
@@
 
-Finally, the Measurements.jl defines how plotting must be performed for an array of measurements, hence, for example a plotting tool that knows the tabular structure of JuliaDB, such as StatPlots.jl can be readily used to plot the table.
+\item the Measurements.jl defines how plotting must be performed for an array of measurements, hence, for example a plotting tool that knows the tabular structure of JuliaDB, such as StatPlots.jl can be readily used to plot the table.
 The plotting automatically shows error bars representing uncertainty.
 
-The Measurement type in the Measurements.jl ~\cite{giordano2016} package represents physical measurements with some uncertainty.
-JuliaDB is unaware of the Measurement datatype, yet it can store, read and write to disk, and perform any database operation while using the math defined on the measurement type.
+\end{enumerate}
+
 In a two-language system, a performant implementation of the measurement datastructure must be done in a low-level language like C.
@@
 
-Here we show benchmarks of JuliaDB against other popular data science
+Here we show benchmarks of JuliaDB against other popular analytical
 databases. Where appropriate, we turned off JuliaDB's distributed
@@
 
-\section{Future directions}
+\section{Conclusion}
 
-1. AxisArrays
-2. Query planning
-3. More support for unstructured data
+Julia turns out to be an excellent language for the implementation as well as the usage of an analytical database.
+Julia's ability to program close to the metal of the computer helps the database implementer, it's rich ecosystem of numerical packages, and ease of use help the user.
+
+As mentioned we leave lazy-evalutation of operations to future work, possibly in more general packages such as Query.jl or DataStreams.jl.
 
@@
 
-Tanmay Mohapatra, David Anthoff, Jacob Quinn
+The authors would like to thank Tanmay Mohapatra who helped in implementing and debugging the distributed infrastructure underlying JuliaDB.
+We thank David Anthoff for help with the CSV ingestion utility in JuliaDB, as well as effort to integrate it with his own widely used set of analytics packages known as the "queryverse", and Jacob Quinn, for his valuable feedback, bug fixes and the WeakRefString type which StringArray is based on.
 

### 2019-09-24 paper.tex
attemp to fix

@@
 We will defer their discussion to the API documentation of JuliaDB \cite{juliadbapi}.
-Read the Sectoin \ref{sec:sort} for more about the implementation of some of these operations.
+Read the Section~\ref{sec:sort} for more about the implementation of some of these operations.
 
@@
 A \texttt{StringArray}'s stores a collection of strings with less memory indirection than a usual array of strings.
-Its memory layout is show in in Figure \ref{fig:stringarray}.
+Its memory layout is show in in Figure~\ref{fig:stringarray}.
 The contents of strings (unicode character codes) are packed into a contiguous buffer of 8-bit unsigend integers.
